Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems

### 2023-03-08 rmd/manuscript.Rmd
update: si document

Prior to the main simulation, we performed a one-patch simulation to explore the predation parameter space ($a_{ij}$ and $h_{ij}$) that allows stable coexistence of the three species without disturbance. In this simulation, we assumed $h_{ij} = h$ to reduce the dimension of parameter space. This analysis revealed that the stable coexistence of the three species is possible with a certain combinations of predation parameters. Within this parameter space, we used three parameter combinations to represent distinctive food webs with varied levels of omnivory: "chain (no omnivory)" [$(a_{BC}, a_{BP}, a_{CP}, h) = (0.5, 0, 0.0025, 0.5)$], "weak omnivory" ($0.5, 0.02, 0.025, 0.5$), and "strong omnivory" ($0.5, 0.04, 0.0025, 0.75$). We used $s = 1$ except for the chain scenario with $s = 0$. Without the loss of generality, we assumed $e_{ij} = e = 1$.

>>>>

Prior to the main simulation, we performed a one-patch simulation to explore the predation parameter space ($a_{ij}$ and $h_{ij}$) that allows stable coexistence of the three species without disturbance. In this simulation, we assumed $h_{ij} = h$ to reduce the dimension of parameter space. This analysis revealed that the stable coexistence of the three species is possible with a certain combinations of predation parameters. Within this parameter space, we used three parameter combinations to represent distinctive food webs with varied levels of omnivory: "chain (no omnivory)" $(a_{BC}, a_{BP}, a_{CP}, h) = (0.5, 0, 0.0025, 0.5)$, "weak omnivory" ($0.5, 0.02, 0.025, 0.5$), and "strong omnivory" ($0.5, 0.04, 0.0025, 0.75$). We used $s = 1$ except for the chain scenario with $s = 0$. Without the loss of generality, we assumed $e_{ij} = e = 1$.

### 2023-03-08 rmd/suppl_info.Rmd
edit
update: si document

---
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    includes:
      before_body: title.tex
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
header-includes:
  \pagenumbering{gobble}
  \usepackage{amsmath}
---

@@

@@

# Sensitivity data
source(here::here("code/analysis_stvy.R"))
load(file = here::here("output/result_main.RData"))
load(file = here::here("output/result_stvy.RData"))

sim_sv <- sim_stvy_result
n_sim_stvy <- unique(sim_sv$n_warmup) + unique(sim_sv$n_burnin) + unique(sim_sv$n_timestep)

```

@@

## Sensitivity analysis

We performed a sensitivity analysis of the trophic community simulation to identify key simulation parameters that strongly affect the relationships between food chain length and ecosystem properties (the number of habitat patches $N_{p}$ and branching probability $P_{b}$. We generated 500 sets of parameter combinations by randomly drawing values of `r length(attributes(fit_sense$fit[[1]]$terms)$term.label)` simulation parameters from uniform distributions (**Table S1**). For each parameter combination, we generated `r max(sim_sv$n_rep)` branching networks with the gradients of ecosystem size ($N_p \sim Unif(10, 150)$) and complexity ($P_b \sim Unif(0.01, 0.99)$). This results in a total of `r nrow(sim_sv)` simulation replicates. We ran `r n_sim_stvy` time steps of trophic community dynamics in each simulation replicate. We obtained temporal means of food chain length for the last `r unique(sim_sv$n_timestep)` time steps. The first `r n_sim_stvy - unique(sim_sv$n_timestep)` time steps were discarded as initialization and burn-in periods.

For each parameter combination, we estimated the Spearman's rank correlation between food chain length and ecosystem properties ($\psi_1$ for $N_{p}$ and $\psi_2$ for $P_{b}$). To examine influences of simulation parameters (**Table S1**), we developed the following regression model taking $\psi_1$ or $\psi_2$ as a response variable $u_n$ (parameter combination $n$):

```{=tex}
\begin{align}
  u_{n} &\sim Normal(\mu_n, \sigma^2)\\
  \mu_n &= \zeta_0 + \sum_{k}{\zeta_k x_{k, n}}
\end{align}
```
where $\zeta_k$ are the intercept and regression coefficients, and $x_{k,n}$ are the simulation parameters drawn from uniform distributions (**Table S1**). Response and explanatory variables were standardized to a mean of zero and a standard deviation of one so that regression coefficients are comparable.

The sensitivity analysis revealed key simulation parameters. For ecosystem size, dispersal distance $\theta$, local disturbance variation $\sigma_{l}$, and the reproductive rate of basal species $r_b$ were influential. However, parameters that influence food web structure were less influential (attack rate, conversion efficiency, and handling time). For ecosystem complexity, dispersal distance $\theta$ and probability $p_d$, local disturbance variation $\sigma_{l}$, and the reproductive rate of basal species $r_b$ were influential. Food web parameters were also important (attack rate, conversion efficiency, and handling time).

Based on the results, we considered ecological scenarios relevant for putative controls of food chain: low and high productivity ($r_b = 4, 8$), and low and high disturbance ($p_m = 0.01, 0.1$). We also investigated the role of omnivory by changing attack rate ($a_{BP}$ and $a_{CP}$) and conversion efficiency ($e_{BP}$ and $e_{CP}$) of two prey items for predator $P$. Specifically, we used the following combinations to represent three levels of omnivory: "Chain", $\{a_{BP}, a_{CP}, e_{BP}, e_{CP}\}=\{$ `r c(min(sim_main_result$a_bp), max(sim_main_result$a_cp), unique(sim_main_result$e_bp)[1], max(sim_main_result$e_cp))` $\}$; "Weak omnivory", $\{a_{BP}, a_{CP}, e_{BP}, e_{CP}\}=\{$ `r c(min(sim_main_result$a_bp), max(sim_main_result$a_cp), unique(sim_main_result$e_bp)[2], max(sim_main_result$e_cp))` $\}$; and "Strong omnivory", $\{a_{BP}, a_{CP}, e_{BP}, e_{CP}\}=\{$ `r c(max(sim_main_result$a_bp), min(sim_main_result$a_cp), unique(sim_main_result$e_bp)[3], min(sim_main_result$e_cp))` $\}$. To simplify simulation setup, we fixed the following parameters: $e_{BC}=$ `r unique(sim_main_result$e_bc)` , $a_{BC}=$ `r unique(sim_main_result$a_bc)`, and $h_{ij}=$ `r unique(sim_main_result$h_bc)`.

\pagebreak

# Tables

## Table S1 Parameter values

Parameter values used for the main and sensitivity analysis.

```{r table_param}

source(here::here("code/table_param.R"))
knitr::kable(df_value, format = "markdown")

```

\pagebreak

```{r table_sensitivity}

source(here::here("code/table_stvy.R"))

j <- 1

## conditional text
c_text <- c("ecosystem size",
            "ecosystem complexity")

table_stvy_chunk <- knit_chunk <- NULL

for(i in 1:length(table_stvy)) {

  table_legend <- paste0("Sensitivity analysis of the ", c_text[i]," effect on food chain length. Parameter estimates of linear regression models are shown. The response variable is the Spearman's rank correlation between food chain length and ", c_text[i],". Response and explanatory variables (i.e., simulation parameters) were standardized (mean = 0, SD = 1) before the analysis.")

  knit_chunk <- paste0("## Table S", i + j, " Sensitivity analysis for ", c_text[i]," \n",
                       table_legend,
                       "\n```{r table_s", i + j,", results='asis'}
                       \n\n
                       table_stvy[[", i,"]]
                       \n\n
                       ```\n",
                       "\n\\pagebreak\n")

  table_stvy_chunk <- c(table_stvy_chunk, knit_chunk)

}

```

`r paste(knit(text = table_stvy_chunk), collapse = '\n')`

\pagebreak

@@

```{r si_figure_fcl}

source(here::here("code/figure_si_fcl.R"))

j <- 0

## conditional text
c_text <- c("ecosystem size",
            "ecosystem complexity")

fig_fcl_chunk <- knit_chunk <- NULL

for(i in 1:length(list_g_fcl)) {

  K <- length(list_g_fcl[[i]])

  for(k in 1:K) {

      figure_legend <- paste0(str_to_sentence(c_text[i])," effect on food chain length and occupancies of constituent species (proportion of habitat patches occupied). Rows are different response variables, and columns are different simulation scenarios. Lines and shades represent loess curves fitted to the simulated data and associated 95% confidence intervals. In this simulation, the following values of parameters were used: $\\theta$ = ", df_param$theta[k], ", $\\sigma_l$ = ", df_param$sd_disturb_lon[k], ", and $\\sigma_h$ = ", df_param$sd_disturb_source[k], ".")

  knit_chunk <- paste0("## Figure S", K * (i - 1) + k + j, " ", str_to_sentence(c_text[i]), " effect on food chain length", " \n",
                       "\n```{r figure_s", K * (i - 1) + k + j,", fig.height = 10, fig.width = 11.5}
                       \n\n
                       list_g_fcl[[", i, "]][[", k, "]]
                       \n\n
                       ```\n",
                       figure_legend,
                       "\n\\pagebreak\n")

  fig_fcl_chunk <- c(fig_fcl_chunk, knit_chunk)

  }

}

```

`r paste(knit(text = fig_fcl_chunk), collapse = '\n')`

>>>>

---
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    includes:
      before_body: title.tex
mainfont: Times New Roman
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
fontsize: 11pt
header-includes:
  \pagenumbering{gobble}
  \usepackage{amsmath}
---

@@

library(tidyverse)

@@

df_sim1 <- readRDS(here::here("output/sim_one_patch.rds"))
n_sim <- df_sim1 %>% 
  mutate(x = n_timestep + n_warmup + n_burnin,
         y = n_warmup,
         z = n_burnin) %>% 
  dplyr::select(x, y, z) %>% 
  distinct() %>% 
  unlist()
```

@@

## One-patch simulation

We performed one-patch simulations to identify parameter combinations that allow stable coexistence of the three species without dispersal and disturbance. Specifically, we varied attack rates $a_{ij}$, handling time $h_{ij}$, intrinsic growth rate of the basal species $r_b$, and carrying capacity $K$. First, we considered `r nrow(dplyr::distinct(df_sim1, r_b, k, a1, h))` discrete combinations of $a_{BC}$, $r_b$, $K$, and $h$ (assumed $h_{ij} = h$). We further crossed `r nrow(dplyr::distinct(df_sim1, a2))` values of $a_{BP}$ (`r min(df_sim1$a2)` -- `r max(df_sim1$a2)` with an equal interval) and $a_{CP}$ (`r min(df_sim1$a3)` -- `r max(df_sim1$a3)`). Consequently, we had a total of `r nrow(dplyr::distinct(df_sim1, r_b, k, a1, a2, a3, h))` combinations for one-patch simulations.

For each parameter combination, we ran 10 replications of `r n_sim[1]` time steps, which include `r n_sim[2]` warmup and `r n_sim[3]` burn-in periods. We recorded the patch state as numeric values ("No species" = 0, "B" = 1, "B + C" = 2, "B + P" =3, and "B + C + P" = 4) at $t = 1600$ and retrieved the maximum value of 10 replicates.

Results were summarized in **Figures S1 and S2**. We identified the following three parameter combinations that could maintain the state of "B + C + P" with distinct food web motifs: "chain (no omnivory)" $(a_{BP}, a_{CP}, h) = (0, 0.0025, 0.5)$, "weak omnivory" ($0.02, 0.025, 0.5$), and "strong omnivory" ($0.04, 0.0025, 0.75$) (see symbols in **Figures S1 and S2**). In the maintext, we reported the results with $a_{BC} = 0.5$ because qualitative patterns did not depend on $a_{BC}$.

\newpage

@@

```{r fig1, fig.cap="Results of one-patch simulation with $a_{BC} = 0.25$. Colors indicate the patch state, and points with different symbols identify parameter combinations used in the main simulation. Panels distinguishh the value of carrying capacity $K$. In both panels, rows represent different values of intrinsic growth rate $r_b$, and columns identify handling time $h$."}
knitr::include_graphics(here::here("figure/figure_si_one_patch_a1_25.pdf"))
```

\newpage

```{r fig2, fig.cap="Results of one-patch simulation with $a_{BC} = 0.50$. Colors indicate the patch state, and points with different symbols identify parameter combinations used in the main simulation. Panels distinguishh the value of carrying capacity $K$. In both panels, rows represent different values of intrinsic growth rate $r_b$, and columns identify handling time $h$."}
knitr::include_graphics(here::here("figure/figure_si_one_patch_a1_50.pdf"))
```

\newpage

```{r fig3, fig.cap="Relationships between food chain length and ecosystem size with low disturbance intensity $\\mu_m = 0.2$. Rows distinguish food web motifs, and columns indicate different dispersal and productivity scenarios."}
knitr::include_graphics(here::here("figure/figure_si_n_patch.pdf"))
```

\newpage

```{r fig4, fig.cap="Relationships between food chain length and ecosystem complexity with low disturbance intensity $\\mu_m = 0.2$. Rows distinguish food web motifs, and columns indicate different dispersal and productivity scenarios."}
knitr::include_graphics(here::here("figure/figure_si_p_branch.pdf"))
```

\newpage

```{r fig5, fig.cap="Relationships between patch state occupancy and ecosystem size with high disturbance intensity $\\mu_m = 0.8$. Panels indicate different food web motifs. In each panel, rows distinguish patch state, and columns indicate different dispersal and productivity scenarios."}
knitr::include_graphics(here::here("figure/figure_si_state_n_patch.pdf"))
```

\newpage

```{r fig6, fig.cap="Relationships between patch state occupancy and ecosystem complexity with high disturbance intensity $\\mu_m = 0.8$. Panels indicate different food web motifs. In each panel, rows distinguish patch state, and columns indicate different dispersal and productivity scenarios."}
knitr::include_graphics(here::here("figure/figure_si_state_p_branch.pdf"))
```

\newpage

```{r fig7, fig.cap="Effects of productivity (left column) and dispersal (right column) on food chain length grouped by disturbance frequency ($p_m$; x-axis). Colors distinguish productivity or dispersal scenarios, and rows identify food web motifs. Disturbance intensity $\\mu_m = 0.8$."}
knitr::include_graphics(here::here("figure/figure_si_fcl_r_disp.pdf"))
```

\newpage

```{r fig8, fig.cap="Effects of productivity on state transitions. Colors distinguish productivity scenarios, and rows identify food web motifs. Disturbance intensity $\\mu_m = 0.8$."}
knitr::include_graphics(here::here("figure/figure_si_state_r.pdf"))
```

\newpage

```{r fig9, fig.cap="Effects of dispersal on state transitions. Colors distinguish distinguish scenarios, and rows identify food web motifs. Disturbance intensity $\\mu_m = 0.8$."}
knitr::include_graphics(here::here("figure/figure_si_state_disp.pdf"))
```
